From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-08-02 WOLFHPC15/article/ms_msc.tex
4.2

%----------------------------------------
\subsection{MSC Parallelization}
%----------------------------------------
The MSC parallelization, which is a subpart of the overall compiler MSCAC, makes use of the ordered list of computations $\Gamma$, which can directly be extracted from the parser, to build a parallel representation of the computations of the overall multi-stencil program. This parallelization phase of the compilation is divided in five different steps; it transforms $\Gamma$ to a series-parallel tree decomposition~\cite{Valdes:1979:RSP:800135.804393}. This transformation phase handles data parallelism through the first step, and task parallelism through the three next steps. This section gives an overview of the five steps of the MSC transformation; the detailed formalism and the associated algorithms of those steps are described in the research report~\cite{????}.

@@

\subsubsection*{The Synchronized Ordered List of Computations: $\Gamma_{sync}$}
Considering a data parallelization of the simulation, the mesh and data of the simulation are splitted among available processors (or cores) and a single program is applied on each sub-part with additionnal synchronizations between processors. For example, as a stencil kernel accesses neighborhood values, values computed by another processor are needed and synchronized. 
From the ordered list of computations $\Gamma$, the needed synchronizations can be automatically added. A synchronization is needed each time a data read by a stencil computation has been written by a previous computation. In such a case, a computation is added before the stencil computation. This \emph{synchronization computation} read the data to synchronize, and write the same data. The computation domain of such a synchronization computation is the mesh entity on which the data is declared. As a result $\Gamma$ is transformed to a synchronized ordered list of computations $\Gamma_{sync}$.
For example, in Figure~\ref{fig:mslex}, the stencil computation $c_1$ read the data $b$ which has been written by the computation $c_0$. For this reason the sublist $[c_0,c_1]$ of $\Gamma$ is transformed to the sublist $[c_0,sync_1,c_1]$ in $\Gamma_{sync}$. The new computation $sync_1$ read and write $b$ and is applied on the computation domain of $c_1$. As a result, a dependency is kept between $c_0$ and $sync_1$ and between $sync_1$, and $c_1$. The same transformation is performed for all stencils: $c_4$ and $c_8$. Thus, $\Gamma_{sync} = [c_0,sync_1,c_1,c_2,c_3,sync_4,c_4,c_5,c_6,c_7,sync_8,c_8]$.

\subsubsection*{The Dependency Graph: $\Gamma_{dep}$}
From the synchronized ordered list of computations $\Gamma_{sync}$, a dependency graph $\Gamma_{dep}$ is built. A dependency exists between two computations (including synchronizations) if and only if a data read has been written by a previous computation in $\Gamma_{sync}$, and if the computation domains intersect. In other words, the only case where a data dependency exists but is not taken into account is when the computation domains do not intersect. Nodes of the dependency graph represent computations of $\Gamma_{sync}$, while edges are dependencies between them. The dependency graph $\Gamma_{dep}$ is a directed acyclic graph (\emph{dag}). For example, the dependency dag $\Gamma_{dep}$ of the example of Figure~\ref{fig:mslex} is displayed in Figure~\ref{fig:hyb}.

@@

\subsubsection*{The Minimal Series-Parallel Graph: $\Gamma_{msp}$}
Once a dependency graph is built, many solutions can be used to build a parallel application, as for example dynamic schedulers~\cite{Augonnet2011,Gautier:2013:XRS:2510661.2511383}. In this work a static scheduling of the dependency graph is built. To do so the dependency graph is transformed to a minimal series-parallel graph~\cite{Valdes:1979:RSP:800135.804393}. As it has been shown in~\cite{Valdes:1979:RSP:800135.804393}, the transitive reduction of a dag is a minimal series-parallel graph if and only if a forbidden shape, called \emph{N-shape} illustrated in Figure~\ref{fig:n}, is not found in the dag. Thus, to transform $\Gamma_{dep}$ to the minimal series-parallel graph $\Gamma_{msp}$, an algorithm~\cite{Mitchell:2004:CMV:1082101.1082117} is applied to remove all \emph{N-shapes} by an over-constraint, as illustrated in Figure~\ref{fig:over}.

>>>>

%----------------------------------------
\subsection{MSC Parallelization}
%----------------------------------------
The MSC parallelization is a subpart of the overall compiler MSCAC.
It makes use of the ordered list of computations $\Gamma$, which can directly be extracted from the parser, to build a parallel representation of the computations of the overall multi-stencil program.
This parallelization phase is divided in five different steps to transform $\Gamma$ to a series-parallel tree decomposition~\cite{Valdes:1979:RSP:800135.804393}.
Data parallelism is handled in the first two step, and task parallelism in the following three steps. This section gives an overview of the five steps of the MSC transformation.
The detailed formalism of these steps and the associated algorithms are described in a research report~\cite{????}.

@@

\subsubsection*{The Synchronized Ordered List of Computations: $\Gamma_{sync}$}
In a data parallelization, data is split among processors and the same program is applied on each sub-part of the data.
This does however require additional synchronizations between processors.
For example, as a stencil kernel accesses neighbor values, values computed by another processor are needed and must be synchronized. 
The required synchronizations can be automatically computed from the ordered list of computations $\Gamma$.
A synchronization is needed each time a data read by a stencil computation has been written by a previous computation. In such a case, a computation is added before the stencil computation. This \emph{synchronization computation} reads the data to synchronize, and write the same data. The computation domain of such a synchronization computation is the mesh entity on which the data is declared. As a result $\Gamma$ is transformed to a synchronized ordered list of computations $\Gamma_{sync}$.
For example, in Figure~\ref{fig:mslex}, the stencil computation $c_1$ read the data $b$ which has been written by the computation $c_0$. For this reason the sublist $[c_0,c_1]$ of $\Gamma$ is transformed to the sublist $[c_0,sync_1,c_1]$ in $\Gamma_{sync}$. The new computation $sync_1$ read and write $b$ and is applied on the computation domain of $c_1$. As a result, a dependency is kept between $c_0$ and $sync_1$ and between $sync_1$, and $c_1$. The same transformation is performed for all stencils: $c_4$ and $c_8$. Thus, $\Gamma_{sync} = [c_0,sync_1,c_1,c_2,c_3,sync_4,c_4,c_5,c_6,c_7,sync_8,c_8]$.

\subsubsection*{The Dependency Graph: $\Gamma_{dep}$}
From the synchronized ordered list of computations $\Gamma_{sync}$, a dependency graph $\Gamma_{dep}$ is built. A dependency exists between two computations (including synchronizations) if and only if a data read has been written by a previous computation in $\Gamma_{sync}$, and if the computation domains intersect. In other words, the only case where the same data is manipulated by distinct kernels without inducing a dependency is when the computation domains do not intersect. Nodes of the dependency graph represent computations of $\Gamma_{sync}$, while edges are dependencies between them. The dependency graph $\Gamma_{dep}$ is a directed acyclic graph (\emph{DAG}). For example, the dependency DAG $\Gamma_{dep}$ of the example of Figure~\ref{fig:mslex} is presented in Figure~\ref{fig:hyb}.

@@

\subsubsection*{The Minimal Series-Parallel Graph: $\Gamma_{msp}$}
Once a dependency graph is built, many solutions can be used to build a parallel application, as for example dynamic schedulers~\cite{Augonnet2011,Gautier:2013:XRS:2510661.2511383}. In this work a static scheduling of the dependency graph is built. To do so the dependency graph is transformed to a minimal series-parallel graph~\cite{Valdes:1979:RSP:800135.804393}. As it has been shown in~\cite{Valdes:1979:RSP:800135.804393}, the transitive reduction of a DAG is a minimal series-parallel graph if and only if a forbidden shape, called \emph{N-shape} illustrated in Figure~\ref{fig:n}, is not found in the DAG. Thus, to transform $\Gamma_{dep}$ to the minimal series-parallel graph $\Gamma_{msp}$, an algorithm~\cite{Mitchell:2004:CMV:1082101.1082117} is applied to remove all \emph{N-shapes} by an over-constraint, as illustrated in Figure~\ref{fig:over}.
